Abstraction and Invariance for Algebraically Indexed Types

### 2012-07-09 paper/geometry-twocolumn.tex
Lots of little typesetting fixes and typos

\begin{example}
  The evaluation of quadratic B\'{e}zier curves (B\'{e}zier curves
  with two endpoints and a single control point) can be expressed
  using the affine combination primitive as follows: % and three steps
  % of De Casteljau's algorithm:
  \begin{displaymath}
    \begin{array}{@{}l}
      \mathrm{quadB\acute{e}zier} : \forall t
      \mathord:\SynTransl{2}.\ \tyPrim{vec}{t} \to \tyPrim{vec}{t} \to
      \tyPrim{vec}{t} \to \tyPrimNm{real} \to \tyPrim{vec}{t}
      \\ \mathrm{quadB\acute{e}zier}\ [t]\ p_0\ p_1\ p_2\ s = \\ \quad
      \mathrm{affComb}\ [t]\ (\mathrm{affComb}\ [t]\ p_0\ s\ p_1)\ s\ (\mathrm{affComb}\ [t]\ p_1\ s\ p_2)
      \\
    \end{array}
  \end{displaymath}
  For two endpoints $p_0$ and $p_2$, a control point $p_1$, and $s \in
  [0,1]$, an application
  $\mathrm{quadB\acute{e}zier}\ p_0\ p_1\ p_2\ s$ gives the point on
  the curve at ``time'' $s$.  The type of
  $\mathrm{quadB\acute{e}zier}$ immediately tells us that it preserves
  all changes of origin.
\end{example}
% Affine combination only provides us with an operation on points. We
% are also interested in operations on the vectors representing offsets
% between points. We now examine the correct types to assign to the
% vector space operations of addition of vectors, negation of vectors,
% multiplication by a scalar and the zero vector. The typings of these
% operations will make use of the abelian group structure of change of
% origin translations.

@@

\begin{example}\label{ex:type-iso}
  The vector space operators and the properties that follow from their
  types allow us to establish a useful type isomorphism. Consider
  functions with types following the schema:
  \begin{displaymath}
    \tau_{n} \isDefinedAs \forall
    t\mathord:\SynTransl{2}.\ \underbrace{\tyPrim{vec}{t} \to ... \to
      \tyPrim{vec}{t}}_{n+1\textrm{ times}} \to \tyPrimNm{real}
  \end{displaymath}
  Just by looking at the types $\tau_{n}$, we know that their
  inhabitants will be invariant under change of origin because of the
  quantification over all $t$ in $\SynTransl{2}$. So we may as well
  choose one of the input points as the origin and assume that all the
  other points are defined with respect to it.  This formalises the
  common mathematical practice of stating that ``without loss of
  generality'' we can take some point in a description of a problem to
  be the origin provided the problem statement is invariant under
  translation. Each type $\tau_{n}$ is isomorphic to the coresponding
  type $\sigma_{n}$:
  \begin{displaymath}
    \sigma_{n} \isDefinedAs \underbrace{\tyPrim{vec}{0} \to
      ... \tyPrim{vec}{0}}_{n\textrm{ times}} \to \tyPrimNm{real}
  \end{displaymath}
  We demonstrate these isomorphisms formally in
  \autoref{sec:instantiations}, in the more general setting of types
  indexed by abelian groups.
  % The isomorphism between these two families of types can be witnessed
  % by the following two functions.
  % \begin{displaymath}
  %   \begin{array}{l}
  %     i_{n,A} : \tau_{n,A} \to \sigma_{n,A} \\
  %     i_{n,A}\ f\ [B]\ (p_1, p_2, ..., p_n) = f\ [B \ltimes 0]\ (0, p_1, ..., p_n) \\
  %     \\
  %     i^{-1}_{n,A} : \sigma_{n,A} \to \tau_{n,A} \\
  %     i^{-1}_{n,A}\ g\ [F]\ (p_0, ..., p_n) = g\ [\pi_L(F)]\ (p_1-p_0, ..., p_n-p_0)
  %   \end{array}
  % \end{displaymath}
  % The direction defined by the function $i_{n,A}$ treats the inputs
  % $p_1,...,p_n$ as vector offsets from the origin $0$. The direction
  % witnessed by $i^{-1}_{n,A}$ picks the first point to act as the
  % origin, and uses the operation of vector subtraction we defined
  % above to turn each of the other points into offsets from this
  % point. Note that this isomorphism is not unique: for each $n$ we can
  % pick any of the $n+1$ inputs to act as the distinguished
  % origin. Thus the families of types $\tau_{n,A}$ and $\sigma_{n,A}$
  % are isomorphic, but not uniquely isomorphic.
\end{example}

\begin{example}\label{ex:uninhabited-type}
  So far we have emphasised the derivation of properties, or ``free
  theorems'', of programs from their types. But using more refined
  relational interpretations of types we can also show that certain
  types are uninhabited. For example, the type $\forall t \mathord:
  \SynTransl{2}.\ \tyPrim{vec}{t + t} \to \tyPrim{vec}{t}$ has no
  quantifier-free inhabitants\footnote{We define quantifier-free
    programs in \thmref{thm:abstraction}.}. Intuitively, this is
  because there is no way to remove the extra $t$ in
  $\tyPrim{vec}{t+t}$ using the vector operations. We formalise this
  indefinability result in \autoref{sec:types-indexed-abelian-groups-indef}
  using a more sophisticated relational interpretation of free
  translation variables than the one we sketched above.
\end{example}

@@

Linear maps of the form $\left(
  \begin{smallmatrix}s & 0 \\ 0 & s\end{smallmatrix}\right)$, as
generated by $\iota_1$, commute with all other invertible linear
maps. We thus require
$(\iota_1 s)B = B(\iota_1 s)$ to hold.
The scaling maps $\left(
  \begin{smallmatrix}s & 0 \\ 0 & s\end{smallmatrix}\right)$
are precisely the elements of $\GL{2}$ that commute with all others;
these form the \emph{centre} of $\GL{2}$. If we keep track of
scalings, then we can assign the following more precise types to
scalar multiplcation and dot product:
\begin{displaymath}
\begin{array}{l@{}l}
  (*) : \forall s \mathord: \SynGL{1}, B \mathord:
  \SynGL{2}.\ & \tyPrim{real}{s} \to \tyPrim{vec}{B,0} \to
  \tyPrim{vec}{\iota_1(s)B,0}\\
%\end{displaymath}
%and dot product:
%\begin{displaymath}
%  \begin{array}{l@{}l}
    (\cdot) : \forall s \mathord: \SynGL{1}, O \mathord: \SynOrth{2}.\ & \tyPrim{vec}{\iota_1(s)\iota_O(O),0} \to\\
    &\tyPrim{vec}{\iota_1(s)\iota_O(O),0} \to \tyPrim{real}{s^2}
  \end{array}
\end{displaymath}

>>>>

\begin{example}
  The evaluation of quadratic B\'{e}zier curves (B\'{e}zier curves
  with two endpoints and a single control point) can be expressed
  using the affine combination primitive as follows: % and three steps
  % of De Casteljau's algorithm:
  \begin{displaymath}
    \begin{array}{@{}l}
      \mathrm{quadB\acute{e}zier} : \forall t
      \mathord:\SynTransl{2}.\ \tyPrim{vec}{t} \mathord\to \tyPrim{vec}{t}\mathord\to
      \tyPrim{vec}{t} \mathord\to \tyPrimNm{real} \mathord\to \tyPrim{vec}{t}
      \\ \mathrm{quadB\acute{e}zier}\ [t]\ p_0\ p_1\ p_2\ s = \\ \quad
      \mathrm{affComb}\ [t]\ (\mathrm{affComb}\ [t]\ p_0\ s\ p_1)\ s\ (\mathrm{affComb}\ [t]\ p_1\ s\ p_2)
      \\
    \end{array}
  \end{displaymath}
  For two endpoints $p_0$ and $p_2$, a control point $p_1$, and $s \in
  [0,1]$, an application
  $\mathrm{quadB\acute{e}zier}\ p_0\ p_1\ p_2\ s$ gives the point on
  the curve at ``time'' $s$.  The type of
  $\mathrm{quadB\acute{e}zier}$ immediately tells us that it preserves
  all changes of origin.
\end{example}
% Affine combination only provides us with an operation on points. We
% are also interested in operations on the vectors representing offsets
% between points. We now examine the correct types to assign to the
% vector space operations of addition of vectors, negation of vectors,
% multiplication by a scalar and the zero vector. The typings of these
% operations will make use of the abelian group structure of change of
% origin translations.

@@

\begin{example}\label{ex:type-iso}
  The vector space operators and the properties that follow from their
  types allow us to establish a useful type isomorphism. Consider
  functions with types following the schema:
  \begin{displaymath}
    \tau_{n} \isDefinedAs \forall
    t\mathord:\SynTransl{2}.\ \underbrace{\tyPrim{vec}{t} \to ... \to
      \tyPrim{vec}{t}}_{n+1\textrm{ times}} \to \tyPrimNm{real}
  \end{displaymath}
  Just by looking at the types $\tau_{n}$, we know that their
  inhabitants will be invariant under change of origin because of the
  quantification over all $t$ in $\SynTransl{2}$. So we may as well
  choose one of the input points as the origin and assume that all the
  other points are defined with respect to it.  This formalises the
  common mathematical practice of stating that ``without loss of
  generality'' we can take some point in a description of a problem to
  be the origin provided the problem statement is invariant under
  translation. Each type $\tau_{n}$ is isomorphic to the corresponding
  type $\sigma_{n}$:
  \begin{displaymath}
    \sigma_{n} \isDefinedAs \underbrace{\tyPrim{vec}{0} \to
      ... \tyPrim{vec}{0}}_{n\textrm{ times}} \to \tyPrimNm{real}
  \end{displaymath}
  We demonstrate these isomorphisms formally in
  \autoref{sec:instantiations}, in the more general setting of types
  indexed by abelian groups.
  % The isomorphism between these two families of types can be witnessed
  % by the following two functions.
  % \begin{displaymath}
  %   \begin{array}{l}
  %     i_{n,A} : \tau_{n,A} \to \sigma_{n,A} \\
  %     i_{n,A}\ f\ [B]\ (p_1, p_2, ..., p_n) = f\ [B \ltimes 0]\ (0, p_1, ..., p_n) \\
  %     \\
  %     i^{-1}_{n,A} : \sigma_{n,A} \to \tau_{n,A} \\
  %     i^{-1}_{n,A}\ g\ [F]\ (p_0, ..., p_n) = g\ [\pi_L(F)]\ (p_1-p_0, ..., p_n-p_0)
  %   \end{array}
  % \end{displaymath}
  % The direction defined by the function $i_{n,A}$ treats the inputs
  % $p_1,...,p_n$ as vector offsets from the origin $0$. The direction
  % witnessed by $i^{-1}_{n,A}$ picks the first point to act as the
  % origin, and uses the operation of vector subtraction we defined
  % above to turn each of the other points into offsets from this
  % point. Note that this isomorphism is not unique: for each $n$ we can
  % pick any of the $n+1$ inputs to act as the distinguished
  % origin. Thus the families of types $\tau_{n,A}$ and $\sigma_{n,A}$
  % are isomorphic, but not uniquely isomorphic.
\end{example}

\begin{example}\label{ex:uninhabited-type}
  So far we have emphasised the derivation of properties, or ``free
  theorems'', of programs from their types. But using more refined
  relational interpretations of types we can also show that certain
  types are uninhabited. For example, the type $\forall t \mathord:
  \SynTransl{2}.\ \tyPrim{vec}{t + t} \to \tyPrim{vec}{t}$ has no
  quantifier-free inhabitants\footnote{We define quantifier-free
    programs in the statement of
    \thmref{thm:abstraction}.}. Intuitively, this is because there is
  no way to remove the extra $t$ in $\tyPrim{vec}{t+t}$ using the
  vector operations. We formalise this indefinability result in
  \autoref{sec:types-indexed-abelian-groups-indef} using a more
  sophisticated relational interpretation of free translation
  variables than the one we sketched above.
\end{example}

@@

Linear maps of the form $\left(
  \begin{smallmatrix}s & 0 \\ 0 & s\end{smallmatrix}\right)$, as
generated by $\iota_1$, commute with all other invertible linear
maps. We thus require
$(\iota_1 s)B = B(\iota_1 s)$ to hold.
The scaling maps $\left(
  \begin{smallmatrix}s & 0 \\ 0 & s\end{smallmatrix}\right)$
are precisely the elements of $\GL{2}$ that commute with all others;
these form the \emph{centre} of $\GL{2}$. If we keep track of
scalings, then we can assign the following more precise types to
scalar multiplcation and dot product:
\begin{displaymath}
\begin{array}{@{}l@{}l}
  (*) : \forall s \mathord: \SynGL{1}, B \mathord:
  \SynGL{2}.\ & \tyPrim{real}{s} \to \tyPrim{vec}{B,0} \to
  \tyPrim{vec}{\iota_1(s)B,0}\\
%\end{displaymath}
%and dot product:
%\begin{displaymath}
%  \begin{array}{l@{}l}
    (\cdot) : \forall s \mathord: \SynGL{1}, O \mathord: \SynOrth{2}.\ & \tyPrim{vec}{\iota_1(s)\iota_O(O),0} \to\\
    &\tyPrim{vec}{\iota_1(s)\iota_O(O),0} \to \tyPrim{real}{s^2}
  \end{array}
\end{displaymath}

### 2012-07-09 paper/section-general-framework.tex
Lots of little typesetting fixes and typos

\begin{example}[Two-Dimensional Geometry]
  \label{ex:two-dim-geo-axioms}
  In \autoref{sec:motivating-examples} we assumed 
  %that 
  various equational axioms %hold 
  for indexing expressions standing for
  elements of geometric groups. Assuming the abelian group axioms for
  translations we can formalise this in our %general 
  framework:
  %We can now make this assumption formal
  %in our general framework. Semantically, translations form an abelian
  %group under addition, so we assume the abelian group axioms for
  %translations:
  \begin{displaymath}
    \begin{array}{l}
      t : \SynTransl{2} \vdash t + 1 \stackrel{ax}\equiv t : \SynTransl{2} \\
      t_1, t_2, t_3 : \SynTransl{2} \vdash t_1 + (t_2 + t_3) \stackrel{ax}\equiv (t_1 + t_2) + t_3 : \SynTransl{2} \\
      t : \SynTransl{2} \vdash t + (-t) \stackrel{ax}\equiv 1 : \SynTransl{2} \\
      t_1, t_2 : \SynTransl{2} \vdash t_1 + t_2 \stackrel{ax}\equiv t_2 + t_1 : \SynTransl{2} \\
    \end{array}
  \end{displaymath}
  Similarly, the sort of scale factors $\SynGL{1}$ forms an abelian
  group under multiplication, and the sorts $\SynGL{2}$ and
  $\SynOrth{2}$ form (non-abelian) multiplicative groups, so we assume
  the appropriate axioms. We also assume that the operations $\iota_O,
  \iota_1, \det$ and $|\cdot|$ are group homomorphisms, and that
  expressions of the form $\iota_1(s)$ commute with group
  multiplication in the sort $\SynGL{2}$. The absolute value of the
  determinant of an orthgonal transformation is always $1$, so we also
  assume $|\det(\iota_OO)| \stackrel{ax}\equiv 1$. Likewise, scaling
  maps have a determined determinant: $\det(\iota_1(s))
  \stackrel{ax}\equiv s \cdot s$.
\end{example}

@@

The defining feature of the index erasure interpretation is that 
semantics of a well-indexed type $\tyPrim{X}{e_1,...,e_n}$ is
determined solely by the primitive type $\tyPrimNm{X}$ and not by the
index expressions $e_1,...,e_n$. %Accordingly, 
We thus assume each
primitive type $\tyPrimNm{X} \in \PrimTypeSet$ is assigned a set
$\tyPrimSem{\tyPrimNm{X}}$ and %We 
extend this assignment to %every
well-indexed types by: % induction on the type structure:
\begin{displaymath}
  \begin{array}{@{\hspace{0em}}c@{\hspace{0em}}c}
    \begin{array}{@{}l}
      \tySem{\tyUnit} \isDefinedAs \{*\} \\
      \tySem{A \tyProduct B} \isDefinedAs \tySem{A} \times \tySem{B} \\
      \tySem{A \tyArr B} \isDefinedAs \tySem{A} \to \tySem{B} \\
    \end{array}
    &
    \begin{array}{l}
      \tySem{A + B} \isDefinedAs \tySem{A} + \tySem{B} \\
      \tySem{\tyPrim{X}{e_1,...,e_n}} \isDefinedAs \tyPrimSem{\tyPrimNm{X}} \\
      \tySem{\forall i\mathord:s. A} \isDefinedAs \tySem{A} \\
      \tySem{\exists i\mathord:s. A} \isDefinedAs \tySem{A}
    \end{array}
  \end{array}
\end{displaymath}
% The interpretation of the unit type is a chosen one element set, and
% the interpretations of the function, tuple and sum types are simply
% the corresponding constructions on sets. 
Note that 
the interpretations of the
universal and existential quantifiers 
do indeed
ignore the indexing. %: the
%interpretation of the type $\forall i\mathord:s.A$ is exactly the
%interpretation of the type $A$, and likewise for $\exists
%i\mathord:s.A$.
We then have:

@@

We use the relational interpretation to define semantic equivalence,
which is a sound approximation of contextual equivalence, as a
consequence of \thmref{thm:abstraction}.
\begin{definition}\label{def:semantic-equality}
  Let $\Gamma_{\mathit{ops}}$ and $\eta_{\mathit{ops}}$ be a context
  of primitive operations and its interpretation as in
  \defref{defn:ctxt-equiv}. Let $\relEnv{E}_{\mathit{ops}}$ be a
  substitutive family of sets of relational environments such that for
  all $\Delta$ and $\rho \in \relEnv{E}_{\mathit{ops}}(\Delta)$,
  $(\eta_{\mathit{ops}}, \eta_{\mathit{ops}}) \in
  \rsem{\Gamma}{\relEnv{E}_{\mathit{ops}}}\rho$.

@@

These relational environments are non-compositional in the sense that
elements of $\relEnv{E}(\Delta,\Delta')$ are not only constructed from
elements of $\relEnv{E}(\Delta)$ and $\relEnv{E}(\Delta')$. As a
consequence of their non-compositionality, these relational
environments will only be \emph{pre-substitutive} in the sense of
\defref{defn:valid-rel-env-family}. Therefore our non-definability
results are subject to the conditions in \thmref{thm:abstraction}.

>>>>

\begin{example}[Two-Dimensional Geometry]
  \label{ex:two-dim-geo-axioms}
  In \autoref{sec:motivating-examples} we assumed 
  %that 
  various equational axioms %hold 
  for indexing expressions standing for
  elements of geometric groups. Assuming the abelian group axioms for
  translations we can formalise this in our %general 
  framework:
  %We can now make this assumption formal
  %in our general framework. Semantically, translations form an abelian
  %group under addition, so we assume the abelian group axioms for
  %translations:
  \begin{displaymath}
    \begin{array}{l}
      t : \SynTransl{2} \vdash t + 0 \stackrel{ax}\equiv t : \SynTransl{2} \\
      t_1, t_2, t_3 : \SynTransl{2} \vdash t_1 + (t_2 + t_3) \stackrel{ax}\equiv (t_1 + t_2) + t_3 : \SynTransl{2} \\
      t : \SynTransl{2} \vdash t + (-t) \stackrel{ax}\equiv 0 : \SynTransl{2} \\
      t_1, t_2 : \SynTransl{2} \vdash t_1 + t_2 \stackrel{ax}\equiv t_2 + t_1 : \SynTransl{2} \\
    \end{array}
  \end{displaymath}
  Similarly, the sort of scale factors $\SynGL{1}$ forms an abelian
  group under multiplication, and the sorts $\SynGL{2}$ and
  $\SynOrth{2}$ form (non-abelian) multiplicative groups, so we assume
  the appropriate axioms. We also assume that the operations $\iota_O,
  \iota_1, \det$ and $|\cdot|$ are group homomorphisms, and that
  expressions of the form $\iota_1(s)$ commute with group
  multiplication in the sort $\SynGL{2}$. The absolute value of the
  determinant of an orthgonal transformation is always $1$, so we also
  assume $|\det(\iota_OO)| \stackrel{ax}\equiv 1$. Likewise, scaling
  maps have a determined determinant: $\det(\iota_1(s))
  \stackrel{ax}\equiv s \cdot s$.
\end{example}

@@

The defining feature of the index erasure interpretation is that 
semantics of a well-indexed type $\tyPrim{X}{e_1,...,e_n}$ is
determined solely by the primitive type $\tyPrimNm{X}$ and not by the
index expressions $e_1,...,e_n$. %Accordingly, 
We thus assume each
primitive type $\tyPrimNm{X} \in \PrimTypeSet$ is assigned a set
$\tyPrimSem{\tyPrimNm{X}}$ and %We 
extend this assignment to %every
well-indexed types by: % induction on the type structure:
\begin{displaymath}
  \begin{array}{@{}c@{\hspace{0em}}c@{}}
    \begin{array}{@{}l}
      \tySem{\tyUnit} \isDefinedAs \{*\} \\
      \tySem{A \tyProduct B} \isDefinedAs \tySem{A} \times \tySem{B} \\
      \tySem{A \tyArr B} \isDefinedAs \tySem{A} \to \tySem{B} \\
    \end{array}
    &
    \begin{array}{l}
      \tySem{A + B} \isDefinedAs \tySem{A} + \tySem{B} \\
      \tySem{\tyPrim{X}{e_1,...,e_n}} \isDefinedAs \tyPrimSem{\tyPrimNm{X}} \\
      \tySem{\forall i\mathord:s. A} \isDefinedAs \tySem{A} \\
      \tySem{\exists i\mathord:s. A} \isDefinedAs \tySem{A}
    \end{array}
  \end{array}
\end{displaymath}
% The interpretation of the unit type is a chosen one element set, and
% the interpretations of the function, tuple and sum types are simply
% the corresponding constructions on sets. 
Note that 
the interpretations of the
universal and existential quantifiers 
do indeed
ignore the indexing. %: the
%interpretation of the type $\forall i\mathord:s.A$ is exactly the
%interpretation of the type $A$, and likewise for $\exists
%i\mathord:s.A$.
We then have:

@@

Semantic equivalence of programs is defined in terms of the relational
interpretation. As a consequence of \thmref{thm:abstraction}, semantic
equivalence is a sound approximation of contextual equivalence.
\begin{definition}\label{def:semantic-equality}
  Let $\Gamma_{\mathit{ops}}$ and $\eta_{\mathit{ops}}$ be a context
  of primitive operations and its interpretation as in
  \defref{defn:ctxt-equiv}. Let $\relEnv{E}_{\mathit{ops}}$ be a
  substitutive family of sets of relational environments such that for
  all $\Delta$ and $\rho \in \relEnv{E}_{\mathit{ops}}(\Delta)$,
  $(\eta_{\mathit{ops}}, \eta_{\mathit{ops}}) \in
  \rsem{\Gamma}{\relEnv{E}_{\mathit{ops}}}\rho$.

@@

These relational environments are non-compositional: elements of
$\relEnv{E}(\Delta,\Delta')$ are not purely constructed from elements
of $\relEnv{E}(\Delta)$ and $\relEnv{E}(\Delta')$. As a consequence,
these relational environments will only be
\emph{pre-substitutive}. Therefore our non-definability results are
subject to the quantifier-free conditions in \thmref{thm:abstraction}.

### 2012-07-09 paper/section-info-flow.tex
Lots of little typesetting fixes and typos

Using algebraically indexed types, we refine Tse and Zdancewic's
translation by replacing each type variable $\alpha_l$ with a
primitive type $\tyPrim{T}{e}$ of representations of the truth of a
logical proposition $e$. The relationships between security levels are
now replaced by logical entailment, so we only have functions of type
$\tyPrim{T}{\phi} \to \tyPrim{T}{\psi}$ when $\phi$ entails
$\psi$. Inspired by the PER interpretation of information flow, we
instantiate our general relational framework to interpret
$\tyPrimNm{T}{e}$ as the identity relation when $e$ is true, and the
empty relation when $e$ is false. In doing so, we can recover the
information flow properties of Abadi et al. and Tse and Zdancewic.

\paragraph{Instantiation of the General Framework}
We assume a single indexing sort $\mathsf{prop}$ and assume the
operations and equations of boolean algebra. Thus we have constants
$\top$, $\bot$ and binary operators $\land$, $\lor$ with the axioms of
a bounded lattice, and a unary complementation operator $\lnot$. We
will use $\phi, \psi$ to stand for index expressions of sort
$\mathsf{prop}$. We use an equational presentation to fit with our
general framework, but note that we can define an order on index
expressions as $\phi \leq \psi$ when $\phi = \phi \land \psi$.

@@

To study information flow, we adapt Tse and Zdancewic's translation of
Abadi et al.'s monadic type to our setting. We define a type
abbreviation $T_\phi A = \tyPrim{T}{\phi} \to A$, where $A$ is a type
and $\phi$ is an expression of sort $\mathsf{prop}$. For every $\phi$,
we can endow the types $T_\phi-$ with the structure of a monad. This
is due to the fact that it is an instance of the ``environment'' (or
``reader'') monad \cite{jones95functional}. We read the types
$T_\phi A$ as data of type $A$ ``protected'' by the principal
$\phi$.

As a consequence of the relational interpretation given above, it
follows that if we have an index expression $\phi$ that is interpreted
as some value other than $\top$ in a relational environment $\rho$,
then for all $x,x' \in \tySem{T_\phi\tyPrimNm{bool}}$, we have $(x,x')
\in \rsem{T_\phi\tyPrimNm{bool}}{\relEnv{E}_\infoflow}{\rho}$. In
terms of the reading above, if a principal is dissatisfied
(i.e.,~$\phi \not= \top$), then data protected by this principal is
indistinguishable from any other data. Thus a program cannot get
access to the exact value. From this and \thmref{thm:abstraction}, we
obtain the following information flow result:

>>>>

Using algebraic indexing, we refine Tse and Zdancewic's translation by
replacing each type variable $\alpha_l$ with a primitive type
$\tyPrim{T}{e}$ of representations of the truth of a logical
proposition $e$. The relationships between security levels are now
replaced by logical entailment, so we only have functions of type
$\tyPrim{T}{\phi} \to \tyPrim{T}{\psi}$ when $\phi$ entails
$\psi$. Inspired by the PER interpretation of information flow, we
instantiate our general relational framework to interpret
$\tyPrim{T}{\phi}$ as the identity relation when $\phi$ is true, and
the empty relation when $\phi$ is false. By doing so, we recover the
information flow properties of Abadi et al. and Tse and Zdancewic.

\paragraph{Instantiation of the General Framework}
We assume a single indexing sort $\mathsf{prop}$ and assume the
operations and equations of boolean algebra. Thus we have constants
$\top$, $\bot$ and binary operators $\land$, $\lor$ with the axioms of
a bounded lattice, and a unary complementation operator $\lnot$. We
will use $\phi, \psi$ to stand for index expressions of sort
$\mathsf{prop}$. We use an equational presentation of boolean algebras
to fit with our general framework, but note that we can define an
order on index expressions as $\phi \leq \psi$ when $\phi = \phi \land
\psi$.

@@

We adapt Tse and Zdancewic's translation of Abadi et al.'s monadic
type to our setting. We define a type abbreviation $T_\phi A =
\tyPrim{T}{\phi} \to A$, where $A$ is a type and $\phi$ is an
expression of sort $\mathsf{prop}$. For every $\phi$, we can endow the
types $T_\phi-$ with the structure of a monad. This is due to the fact
that it is an instance of the ``environment'' (or ``reader'') monad
\cite{jones95functional}. We read the types $T_\phi A$ as data of type
$A$ ``protected'' by the principal $\phi$.

As a consequence of the relational interpretation given above, it
follows that if we have an index expression $\phi$ that is interpreted
as some value other than $\top$ in a relational environment $\rho$,
then for all $x,x' \in \tySem{T_\phi\tyPrimNm{bool}}$, we have $(x,x')
\in \rsem{T_\phi\tyPrimNm{bool}}{\relEnv{E}_\infoflow}{\rho}$. Thus if
a principal is dissatisfied (i.e.,~$\phi \not= \top$), then data
protected by this principal is indistinguishable from any other
data. Thus a program cannot get access to the exact value. From this
and \thmref{thm:abstraction}, we obtain the following information flow
result:

### 2012-07-09 paper/section-metric.tex
Lots of little typesetting fixes and typos

The previous two sections have presented examples where the relational
interpretations of primitive types relates a pair of elements by some
transformation if applying the transformation to the first element
yields the second. Thus, the free theorems that we derived directly
take the form of ``invariance'' properties, where some equation holds
between two terms. In this section, we examine another instantiation
of our general framework that relates values when they are within a
certain distance. The free theorems that we obtain inform us of the
effect that programs have on the distances between values. For
example, a program $M$ of type:
\begin{displaymath}
  \forall \epsilon_1,\epsilon_2\mathord:\metricSort.\ \tyPrim{real}{\epsilon_1} \to \tyPrim{real}{\epsilon_2} \to \tyPrim{real}{\epsilon_1 + \epsilon_2}
\end{displaymath}
must satisfy the property that for all $\epsilon_1, \epsilon_2 > 0$
and $x, x', y, y' \in \mathbb{R}$,
\begin{displaymath} % FIXME: include \eta_\Metric
  \begin{array}{l}
    \textrm{if }|x - x'| < \epsilon_1 \textrm{ and } |y - y'| < \epsilon_2 \textrm{ then}\\
    \hspace{2em}|\tmSem{M}\ x\ y - \tmSem{M}\ x'\ y'| < \epsilon_1 + \epsilon_2
  \end{array}
\end{displaymath}
A type system with a relational interpretation that tracks distances
between values has %previously 
been investigated by Reed and Pierce
\cite{reed10distance} in the setting of differential privacy. Their
system uses a linear type discipline to ensure that all programs are
non-expansive (i.e., the distance between the outputs is no more greater
than the distance between the inputs). Here, we instantiate our general
framework %from \autoref{sec:a-general-framework} %in order 
to express
non-expansivity, and more, via algebraically indexed types.

>>>>

\autoref{sec:instantiations} and \autoref{sec:monoid-indexed-types}
presented examples where the relational interpretations of primitive
types relates a pair of elements by some transformation if applying
the transformation to the first element yields the second. Thus, the
free theorems that we derived directly take the form of ``invariance''
properties, where some equation holds between two terms. In this
section, we examine another instantiation of our general framework
that relates values when they are within a certain distance. The free
theorems that we obtain inform us of the effect that programs have on
the distances between values. For example, a program $M$ of type:
\begin{displaymath}
  \forall \epsilon_1,\epsilon_2\mathord:\metricSort.\ \tyPrim{real}{\epsilon_1} \to \tyPrim{real}{\epsilon_2} \to \tyPrim{real}{\epsilon_1 + \epsilon_2}
\end{displaymath}
must satisfy the property that for all $\epsilon_1, \epsilon_2 > 0$
and $x, x', y, y' \in \mathbb{R}$,
\begin{displaymath} % FIXME: include \eta_\Metric
  \begin{array}{l}
    \textrm{if }|x - x'| < \epsilon_1 \textrm{ and } |y - y'| < \epsilon_2 \textrm{ then}\\
    \hspace{2em}|\tmSem{M}\ x\ y - \tmSem{M}\ x'\ y'| < \epsilon_1 + \epsilon_2
  \end{array}
\end{displaymath}
A type system with a relational interpretation that tracks distances
between values has %previously 
been investigated by Reed and Pierce
\cite{reed10distance} in the setting of differential privacy. Their
system uses a linear type discipline to ensure that all programs are
non-expansive (i.e., the distance between the outputs is no more greater
than the distance between the inputs). Here, we instantiate our general
framework %from \autoref{sec:a-general-framework} %in order 
to express
non-expansivity, and more, via algebraically indexed types.

### 2012-07-09 paper/section-monoid.tex
Lots of little typesetting fixes and typos

\begin{theorem}
  \label{thm:monoid-indefinability}
  Let $m$ and $n$ be natural numbers. For all $n$-by-$m$-matricies of
  natural numbers $A$ and $m$-by-$1$ matricies of natural numbers $b$,
  there exists an $M$ in the quantifier-free fragment, with
  \begin{displaymath}
    i_1,\dots,i_m\mathord:\mathsf{M}; \Gamma_\Mon \vdash M : \tyPrim{M}{e_1} \to \dots \to \tyPrim{M}{e_n} \to \tyPrim{M}{e}
  \end{displaymath}
  where $e_j = i_1^{A_{j1}}\dots i_m^{A_{jm}}$ and $e = i_1^{b_1}\dots
  i_m^{b_m}$ iff there is a solution $\vec{x} = (x_1,...,x_n)$ in the
  natural numbers to $A x^\top = b$.
\end{theorem}

>>>>

\begin{theorem}
  \label{thm:monoid-indefinability}
  Let $m$, $n$ be natural numbers. For all $n$-by-$m$-matrices of
  natural numbers $A$ and $m$-by-$1$ matrices of natural numbers $b$,
  there is a quantifier-free program $N$, with
  \begin{displaymath}
    i_1,\dots,i_m\mathord:\mathsf{M}; \Gamma_\Mon \vdash N : \tyPrim{M}{e_1} \to \dots \to \tyPrim{M}{e_n} \to \tyPrim{M}{e}
  \end{displaymath}
  where $e_j = i_1^{A_{j1}}\dots i_m^{A_{jm}}$ and $e = i_1^{b_1}\dots
  i_m^{b_m}$ iff there is a solution $\vec{x} = (x_1,...,x_n)$ in the
  natural numbers to $A x^\top = b$.
\end{theorem}
